Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-04 paper/main.tex
Updates

\subsection{Measures of Conditional Association}
Our proposed method for iterative DAG construction relies on a measure of
conditional association also known as partial association between variables. We
are interested in measuring the association between variables $ X $ and $ Y $
when conditioned on a set of variables $ \bm{Z} $ which is potentially an empty
set, $ \bm{Z} = \emptyset $. Many special case measures of conditional
association have been used in literature depending on the type of $ X $, $ Y $,
and $ \bm{Z} $. Typically, these are the effect size measures of the
conditional independence test $ X \ci Y \rvert \bm{Z} $. In this section, we
give an overview of some of the commonly used measures for different data
types.

\subsubsection{Both $ X $ and $ Y $ are continuous}
In the case when both $ X $ and $ Y $ are continuous variables with the
assumptions of normality and linearly relationship, typically Pearson's
correction coefficients are used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient can be directly computed. When $ \bm{Z} \neq \emptyset
$, partial correlations are used. One of the ways to estimate it is to use two
linear regression models: $ E_X: x \sim \bm{z} $, and $ E_Y: y \sim \bm{z} $.
Then we take the compute the Pearson's correlation coefficient between the
residuals of these two regressions: $ R_X = x - E_X(\bm{z}) $, $ R_Y = y -
E_Y(\bm{z}) $.

@@

\section{Connection to Score Based Methods}
\begin{itemize}
	\item The process of iteratively adding and removing edges in very similar to
		Greedy Equivalence Search (GES).
	\item But not decomposible like scoring methods. 
	\item But Because they make only local changes, can get stuck in local minimas.
	\item Both approach looks for modifications globally.
	\item The sum of unexplained effect provides an absolute measure of model fit 
		unlike scoring methods.
\end{itemize}

>>>>

\subsection{Measures of Conditional Association}
Our proposed method for iterative DAG construction relies on a measure of
conditional association also known as partial association between variables. We
are interested in measuring the association between variables $ X $ and $ Y $
when conditioned on a set of variables $ \bm{Z} $ which is potentially an empty
set, $ \bm{Z} = \emptyset $. Many special case measures of conditional
association have been used in literature depending on the type of $ X $, $ Y $,
and $ \bm{Z} $. Typically, these are the effect size measures of the CI test $
X \ci Y \rvert \bm{Z} $. In this section, we give an overview of some of the
commonly used measures for different data types.

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used.  In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is:

$$ \rho_{X, Y} = \frac{cov(X, Y)}{\sigma_X \sigma_Y} $$

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient can be used. 
One of the ways to estimate it is to use two linear regression models:

\begin{equation}
	\begin{split}
		E_X &: X \sim \bm{Z} \\
		E_Y &: Y \sim \bm{Z} \\
		\rho_{X, Y; \bm{Z}} &= \frac{cov((X - E_X), (Y - E_Y))}{\sigma_{(X - E_X)} \sigma_{(Y-E_Y)}}
	\end{split}
\end{equation}

$ E_X: x \sim \bm{z} $,
and $ E_Y: y \sim \bm{z} $. Then we take the compute the Pearson's correlation
coefficient between the residuals of these two regressions: $ R_X = x -
E_X(\bm{z}) $, $ R_Y = y - E_Y(\bm{z}) $.

@@

\subsection{Connection to Score Based Methods}
\begin{itemize}
	\item The process of iteratively adding and removing edges in very similar to
		Greedy Equivalence Search (GES).
	\item But not decomposible like scoring methods. 
	\item But Because they make only local changes, can get stuck in local minimas.
	\item Both approach looks for modifications globally.
	\item The sum of unexplained effect provides an absolute measure of model fit 
		unlike scoring methods.
\end{itemize}
